---
title: "Interaction Effects of Supplements and Dosage on Tooth Growth in Guniea Pigs"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Data Preparation & Setup

### Libraries
```{r}
#Data Wrangling & Visualization
library(MedDataSets)
library(tidyverse) # includes dplyr, ggplot2, readr, tidyr, purrr, stringr

#Modeling & Tidying Model Output
library(broom) # for augment(), tidy(), and glance()

#ANOVA Diagnostics & Post-Hoc Analysis
library(car) # for Type II/III ANOVA tables & Levene's test
library(emmeans) # for estimated marginal means & post-hoc pairwise contrasts

#Plot Composition
library(patchwork) # for combining ggplot2 figures (+ operator)
library(kableExtra)
```

## Preliminary Exploratory Analysis

Apply the within() method convert the predictor variables, supplement and dosage to factor variables. Check the balance of the data.

```{r}
toothgrowth <- within(ToothGrowth_df, {fsupp = as.factor(supp); 
               fdose = as.factor(dose)})

#Check balance
toothgrowth |>
  group_by(fsupp, fdose) |>
  count()
```

[Discussion:]{.underline} The data is balanced, however the predictor variable dosage, consists of three levels that are unequally spaced, polynomial terms will be added to the model.

### Strip Chart

```{r}
stripchart(len ~ fdose * fsupp, data = toothgrowth, vertical = TRUE,
           col = c("red", "blue", "green"),
           main = "Treatment Effect on Tooth Length in Guinea Pigs",
           xlab = "Dose Level",
           ylab = "Tooth Length")
```

### Boxplot

```{r}
#Boxplot
ggplot(toothgrowth, aes(x = fdose, y = len, fill = fsupp)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(aes(shape = fdose), width = 0.5, size = 3) +
  labs(
    title = "Treatment Effect on Tooth Length in Guinea Pigs",
    x = "Dose Level",
    y = "Tooth Length") +
  theme_minimal()+
  theme(
    plot.title = element_text(hjust = 0.5))
```

### Group Means

```{r}
ms_tooth <- aggregate(len ~ fsupp * fdose, data = toothgrowth, FUN = mean)
```

Plot group means.
```{r}
ggplot(ms_tooth, aes(x = fdose, y = len, col = fsupp)) +
  geom_jitter(aes(shape = fdose), width = 0.5, size = 4) +
  scale_y_continuous(limits = c(0,30)) +
  labs(
    title = "Treatment Effect on Tooth Length in Guinea Pigs",
    x = "Dose Level",
    y = "Tooth Length") +
  theme_minimal()+
  theme(
    plot.title = element_text(hjust = 0.5))
```

[Discussion:]{.underline} Exploratory analysis indicates the level of dosage with each supplement increases tooth length with orange juice on average producing a higher length of tooth growth with the exception of the 2mg dosage, 
vitamin C is slightly higher. 

### Interactions Plot

```{r}
#Interaction plot

with(toothgrowth, interaction.plot(x.factor = fdose, trace.factor = fsupp,
response = len, col = c("black", "red", "blue"), pch = c(18,19,20), type = "b"))

```

[Discussion:]{.underline} Interactions plot points to interactions between orange juice and vitamin C at the 2.0mg dosage level.This suggests supplement 
is influenced by dosage level at 2.0mg, with likely minimal influence at the lower dosage levels. 

## Statistical Modeling & Selection

### Hypothesis

-   $H_0: \alpha \beta_{11} = \alpha\beta_{12} = ... = \alpha\beta_{ab} = 0$. There is no interaction effect between dose and supplements.

-   $H_A:$ some $\alpha\beta's$ are not zero. There are interaction effects between dose and supplement.


### Model Fitting & Comparison
```{r}
p0 <- aov(len ~ fsupp, data = toothgrowth)
p1 <- aov(len ~ fsupp * dose, data = toothgrowth)
p2 <- aov(len ~ fsupp * (dose + I(dose^2)) ,data = toothgrowth)

aov_tooth <- aov(len ~ fsupp * fdose,
             data = toothgrowth)
anova_tooth <- anova(aov_tooth)


p0_tab <- anova(p0)
p1_tab <- anova(p1)
p2_tab <- anova(p2)
```


### Selection Criteria

**AIC Comparisons**
```{r}
aic_tooth <- AIC(p0, p1, p2)

kable(
  aic_tooth,
  caption = "AIC Comparison"
)
```
  
**ANOVA Table: Model Terms Comparison**
```{r}
options(knitr.kable.NA = ' ')
kable(p2_tab,
      caption = "Model Terms Comparison")
```

[Discussion:]{.underline} In Table 1 it can be shown the polynomial interactions model (p2) yielded the lowest AIC value, supporting model with interactions and a quadratic term. However, two–way ANOVA results of the models indicate the interaction term supplement and the quadratic term dosage has no significant effect on the model (see Table 2). Based on the hierarchical principal of 
the two–way ANOVA test, its addition to the final model will be removed.  

**Final Model**

The estimated final model includes the main effects dosage and supplement, the 
interaction between dosage and supplement, and the quadratic dosage term. 

```{r, echo=TRUE}
#Interaction model
#Fit the model
final_tooth <- aov(len ~ fsupp + dose + I(dose^2)
               + fsupp:dose, data = toothgrowth)
```

**Final Model ANOVA Table**
```{r}
final_anova <- anova(final_tooth)

#ANOVA table
options(knitr.kable.NA = ' ')
kable(final_anova,
      caption = "ANOVA Results")
```

## Model Diagnostics & Assumption Testing

A residuals versus fitted values plot. And a normal Q - Q plot with Shapiro - Wilk test.

```{r model-1-diagnostics, fig.show = "hold", out.width = "50%", dpi = 100, fig.cap = "Diagnostic Plots for the Full Additive Model I"}

#Augment the model to get a tidy data frame with residuals (.resid) & fitted values (.fitted)
aug_tooth <- augment(final_tooth)

#Residuals vs Fitted Plot
p_res <- ggplot(aug_tooth, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(se = FALSE, color = "blue", method = "loess", formula = y ~ x) +
  labs(
    title = "Final Model: Residuals vs Fitted",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 12)

#Normal Q-Q Plot
p_qq <- ggplot(aug_tooth, aes(sample = .resid)) +
  stat_qq(alpha = 0.6) +
  stat_qq_line(color = "red") +
  labs(
    title = "Final Model: Normal Q-Q",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 12)

#Combine side-by-side & save cleanly with ggsave()
final_diag <- p_res + p_qq
final_diag
ggsave(
  filename = "../figures/final_diagnostics.png",
  plot = final_diag,
  width = 10,
  height = 4.5,
  dpi = 100
)

#Shapiro - Wilk Test
shap1 <- shapiro.test(resid(final_tooth))
shap1
``` 

[Discussion:]{.underline} The normal Q–Q plot indicates the residuals approximately follow a normal distribution.There is slight deviation from the line and at the tails. To confirm normality the Shapiro–Wilk test is conducted and has resulted in a high W = 0.9828 and p–value of 0.5576, satisfying the 
normality assumption. The residuals versus fitted values indicates equal variance and thus is satisfied, no clear patterns are observed.

## Post-Hoc Analysis & conclusion

### Estimated Marginal Means

```{r, echo=TRUE}
em_tooth <- emmeans(aov_tooth, pairwise ~ fsupp | fdose)
```

### Pairwise Contrast Comparisons
```{r}
em_tb <- em_tooth$contrasts

#Table
kable(em_tb,
      caption = "Pairwise Comparisons",
      digits = 4)
```

### Contrast Visualizations
```{r}
plot(em_tb)
```

[Discussion:]{.underline} To conclude the analysis, a pairwise comparison test is performed to examine the effects of supplement and dosage, and their levels. Both results for the supplements orange juice and vitamin C at the dosage levels 0.5mg and 1.0 yield estimate results of 5.25 (0.5mg) and 5.93 (1.0mg) with statistically significant p–values (see Table 4). Based on these results, we can conclude that orange juice on average yields a higher response on tooth growth in guinea pigs at dosage levels 0.5mg and 1.0mg. For the dosage level 2.0mg, the estimate value was 0.08, and a p–value of 0.9609 indicated in Table 4. We can deduce vitamin C at the 2.0mg dosage level yields a minimally higher response on tooth growth in guinea pigs, nevertheless given the p – value is not statistically significant, the difference between orange juice and vitamin C at this level is negligible.





















